### 2017-04-19 paper/sexSelectionSpacing-ver1.tex
Minor edits

\title{Son Preference, Sex-Selective Abortions, and Birth Spacing
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to three anonymous referees, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of \AA{}rhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

@@

[ways to get at son preference:
- stated fertility preferences
- longer spacing after son
- differential stopping behavior
- biased sex ratio (overall before sex selection, and specifically at birth later)]

[why do these work less well after introduction of sex selection?]

[how can my method remedy those problems?]

[main results]

The extent of son preference has been an important research area for 
decades and is as salient as ever.
Three approaches have been in widespread use in this literature:
birth spacing, parity progression, and sex ratio \citep{Leung1988}.
Before prenatal sex determination became widely available, parents' only 
option if they wanted a son was to try for another pregnancy as soon as 
possible after the birth of a girl.
A preference for sons can therefore lead to shorter spacing until next 
birth after the birth of a daughter than after the birth of a son 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997}.
The resulting shorter spacing is thought to be associated with worse 
health outcomes for the girls 
\citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter, resulting in a lower parity progression
with the number of sons \citep{ben-porath76b,Das1987,Arnold1997,clark00}. 
Finally, increase in the sex ratio---the ration of boys to
girls---can serve as an indication of son preference because 
of differential mortality or underreporting \citep{sen90,sen92,Merli2000,anderson08}.

 and/or use of prenatal sex selection

or, after the introduction of prenatal sex determination, at birth.
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors 
\citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an 
association with use of sex-selective abortions \citep{retherford03b,jha06,abrevaya09}.

India, for example, has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period \citep{Yi1993,park95}.
}

Recent research suggests, however, that son preference in India, when measured 
as ideally having more boys than girls, is decreasing over time and with 
higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}

Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask 
about use of sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have relied on a simple method for 
establishing factors that affect sex selection: use the sex of children born as 
the dependent variable, and estimate the effects of variables on the probability 
of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short
and used the shorter birth spacing after the birth of a girl than after the
birth of a boy as a measure of the degree of son preference.

I address an important questions that the prior literature has been unable
to examine because of the simple method's limitations:
how does birth spacing interact with use of sex-selective abortions?
I introduce a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the duration between 
births.
I use the method to show how spacing between births play an important role in our 
understanding of sex selection decisions.

% [literature on birth spacing - why spacing is important]

What has not previously been appreciated is that the introduction of sex-selective 
abortions substantially changed the relationship between son preference and birth spacing.
The change happens because each abortion significantly increases the duration until the
next birth.
The increase in duration can be divided into three parts starting from the time of
the abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Secondly, the average expected time to conception is about 6 months.
Finally, sex determination tests are generally reliable only from 3 months of gestation onwards.
The result is that each abortion delays the next birth by approximately a year.%
\footnote{
There are three well-developed technologies for determining a fetus' sex: 
Chorionic villus sampling (CVS), amniocentesis, and ultrasound.
CVS can be applied after the shortest period
of gestation (8 to 12 weeks).
It is the most complicated, but also the most reliable, and an abortion can be done in 
the first trimester.
Amniocentesis can be performed after fourteen weeks, but requires three to four weeks
before the result is available, so an abortion cannot be performed until more than 
halfway through the second trimester.
Ultrasound has the advantages of being non-invasive and relatively cheap but 
is associated with higher risk of faulty sex determination if done early.
Generally, a fetus' sex can be determined in the third month of gestation if it is a
boy and the fourth month if it a girl.

The waiting time to conception does vary by woman, but
even if it is very short, say one  month, the minimum additional space between
births would be 6 months per abortion.
For a discussion of time to conception see \citep{Wang2003}
}

% [why this matters]
How spacing interacts with sex selection is important for two reasons.
First, 
spacing, by itself, can no longer be used to predict son preference. 
Because each abortions extends spacing by about a year, we now have a situation 
where the \emph{longest} spacing is likely observed for families with the 
strongest son preference.
To further complicate matters, 
we may still observe short spacing after the birth of daughter as a representation 
of son preference for families who either have less access to sex-selective 
abortions or are willing to have more children.
Spacing is still important and useful in understanding sex selection and son
preference, but only if combined with the outcome of that spacing---in other words 
the likelihood of observing a boy or a girl.

% using spacing itself to predict son preference becomes useless if
% families have low desired fertility and access to sex-selective abortions!!!!
Second, families may reverse their decision to use sex selection in-between births.
A reversal could happen if they do not want the space between children to be 
``too long'', or if there are concerns about possible infertility as a result of 
too many abortions in a row without a birth.
Hence, for a given parity, the sample of women for whom we do observe a birth 
may behave differently from the sample of women who have not yet had a birth.
This is a problem because births to women with shorter spacing are more likely 
to be in the survey for a given parity.
Hence, if decisions on sex selection change with duration since last birth,
the predicted sex ratio for a parity using the simple method will be a 
biased estimate of the final observed sex ratio for the parity when 
childbearing is complete.

% [contribution of paper]
[TK maybe not focus so much on the simple method but rather understanding son preference].
The main contribution of this paper is the introduction and application of a method that 
incorporates the decisions about birth spacing, and sex-selective abortions.
The empirical model is a competing risk, non-proportional hazard model with two exit 
states: either a boy or a girl is born.
This approach has TK major advantages over the simple method

@@

% [EXPLAIN RESULTS HERE]
% Main findings!
I apply the method to birth histories from India's National Family and Health Surveys 
for Hindu women covering the period 1972 to 2006. 
There are three main findings.

@@

\citep{Soest2012} for a recent example of spacing and son preference

\section{The Interaction between Son Preference and Birth Spacing\label{sec:model}}

With the introduction of sex-selective abortions, parents can now abort
a fetus of unwanted sex.
With an approximately 12 months additional duration per abortion,
this can result in substantially longer durations between births, and
``too long'' spacing may become a greater concern for parents than 
``too short'' spacing.%
\footnote{
Shorter spacing was likely less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
}
If the duration from last birth becomes long enough, parents may 
reverse their decision to use prenatal sex determination
and carry the next pregnancy to term whether male or female.

A number of factors make a reversal of the decision to use sex 
selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that any 
child is better than not having an additional child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although the latter two factor are most often associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

For a given sex composition of prior children, there are two main drivers 
of sex selection: the degree of son preference and the cost of sex selection.
The more parents care about having a son as their next child, the more
likely they are to use sex selection and the less likely they are to
reverse the decision once started.
The result is an increased number of abortions, more spells where a
prenatal scan is used, and the spacing between births is longer.

Lowering the cost of sex selection bias the overall sex ratio further 
towards boys.
Correspondingly, both the number of abortions per woman and the
number of spells where a prenatal scan is used should also increase,
resulting in longer average spacing between births.
This increase in spacing is, however, not likely to be uniform across parities.
In addition to increasing the overall sex ratio, lower cost of prenatal 
scan should also lead to use of sex selection at earlier parities.
It is therefore possible to see either no change or a \emph{decrease} 
in average spacing for a later parity---although with a more than compensating 
increase in an earlier parity.
In other words, it is possible for individual parities to show a \emph{decline}
in sex selection at the same time as the overall use goes up.
This is an important caution against over-interpreting individual 
parity results.

Lowering the cost of prenatal sex determination generally leads to 
a smaller percentage of parents who change their use of sex selection.
But, the total number of spells with a change may still
go up because lower cost of prenatal sex determination also
increases the use of sex selection.

If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method [TK explain what they are instead!].

[TK find examples of the son preference spacing papers and look at what they did
for "theory"]

% What are the model highlights 
There are three main conclusions from the model.
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

% Empirical method
The model highlights two major limitations of the simple 
empirical method, which a better empirical approach needs to 
allow for.
First, increased use of sex selection leads to longer spacing which,
everything else equal, reduces the number of births available for 
estimation in the simple model and thereby lowers its precision
because fewer parents will make it to a given parity by the time of the survey.
Furthermore, the more intensively sex selection is used the larger 
the decline in sample size.
Second, by using only observed births the simple method cannot capture
when parents change their decision on the use of sex selection;
with a reversal the spell can still result in the birth of a girl.
This last point is especially important when trying to establish whether 
son preference has changed over time in a situation where the cost of prenatal 
sex determination has likely \emph{increased} over time because of the PNDT act.
As I show an increase the cost of prenatal sex determination does 
decrease the sex ratio and lower the number of abortions, but 
it also \emph{increases} the likelihood that parents change their initial
decision to use sex selection within a spell.

>>>>

\title{Son Preference, Sex-Selective Abortions,\\ and Birth Spacing
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to three anonymous referees, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of \AA{}rhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

@@

% [ways to get at son preference:
% - stated fertility preferences
% - longer spacing after son
% - differential stopping behavior
% - biased sex ratio (overall before sex selection, and specifically at birth later)]
% 
% [why do these work less well after introduction of sex selection?]
% 
% [how can my method remedy those problems?]
% 
% [main results]

Understanding and measuring son preference has been an important research 
area for decades and is as salient as ever.
[brief description of where identified]
Three approaches to measuring son preference have been in widespread use:
birth spacing, parity progression, and sex ratio \citep{Leung1988}.

Before prenatal sex determination became widely available, parents' only 
option if they wanted a son was to try for another pregnancy as soon as 
possible after the birth of a girl.
A preference for sons is therefore often associated with shorter birth spacing 
after the birth of a daughter than after the birth of a son 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012}.
The resulting shorter spacing is thought to be associated with worse 
health outcomes for the girls 
\citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter, resulting in a lower parity progression
with the number of sons \citep{ben-porath76b,Das1987,Arnold1997,clark00}. 
Finally, increase in the ratio of males to females 
can serve as an indication of son preference because 
of differential mortality, for example because of the shorter
birth spacing, or underreporting \citep{sen90,sen92,Merli2000,anderson08}.

After the introduction of prenatal sex determination, a substantial
literature has focused on the use of prenatal sex selection as a direct 
measure of son preference.
Our understanding of sex selection usage is, however, constrained by a
lack of information;
there are no official data on sex selection, and the few surveys that ask 
about use of sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have instead used the sex of children 
born as the dependent variable, and estimate the effects of variables on the 
probability of having a son.
The idea is that in the absence of any interventions, the probability of having 
a son is approximately 0.512, and this is independent of genetic 
factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant coefficient therefore 
indicates an association between the variable and use of sex-selective 
abortions.
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

What has not previously been appreciated, however, is that the introduction 
of sex-selective abortions substantially changed the relationship between 
son preference and birth spacing.
The change happened because each abortion significantly increases the duration 
until the next birth.
This increase in duration can be divided into three parts starting from the time of
the abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Secondly, the average expected time to conception is about 6 months.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.%
\footnote{
There are three well-developed technologies for determining a fetus' sex: 
Chorionic villus sampling (CVS), amniocentesis, and ultrasound.
CVS can be applied after the shortest period
of gestation (8 to 12 weeks).
It is the most complicated, but also the most reliable, and an abortion can be done in 
the first trimester.
Amniocentesis can be performed after fourteen weeks, but requires three to four weeks
before the result is available, so an abortion cannot be performed until more than 
halfway through the second trimester.
Ultrasound has the advantages of being non-invasive and relatively cheap but 
is associated with higher risk of faulty sex determination if done early.
Generally, a fetus' sex can be determined in the third month of gestation if it is a
boy and the fourth month if it a girl.
}
The result is that each abortion delays the next birth by approximately 
a year.%
\footnote{
The waiting time to conception does vary by woman, but
even if it is very short, say one  month, the minimum additional space between
births would be 6 months per abortion.
For a discussion of time to conception see \citep{Wang2003}
}

% [why this matters]
How spacing interacts with sex selection is important for two reasons.
First, spacing, by itself, can no longer be used to predict son preference. 
Because each abortion extends spacing by about a year, we now have a situation 
where the \emph{longest} spacing is likely observed for families with the 
strongest son preference.
To further complicate matters, 
we may still observe short spacing after the birth of daughter as a representation 
of son preference for families who either have less access to sex-selective 
abortions or are willing to have more children.
Spacing is still important and useful in understanding sex selection and son
preference, but only if combined with the outcome of that spacing---in other words 
the likelihood of observing a boy or a girl.

% using spacing itself to predict son preference becomes useless if
% families have low desired fertility and access to sex-selective abortions!!!!
Second, if the duration from last birth becomes long enough, parents may 
reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
Hence, for a given parity, the sample of women for whom we do observe a birth 
may behave differently from the sample of women who have not yet had a birth.
This is a problem because births to women with shorter spacing are more likely 
to be in the survey for a given parity.
Hence, if decisions on sex selection change with duration since last birth,
the predicted sex ratio for a parity using the current method will be a 
biased estimate of the final observed sex ratio for the parity when 
childbearing is complete.

A number of factors make a reversal of the decision to use sex 
selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that a 
child of any sex is better than not having a child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although the latter two factor are most often associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

% [contribution of paper]
% [TK maybe not focus so much on the simple method but rather understanding son preference].
My main contribution here is the introduction and application
of a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the 
duration between births.
[TK move to estimation strategy?]
The empirical model is a competing risk, non-proportional hazard model with 
two exit states: either a boy or a girl is born.
This approach has TK major advantages over the simple method

@@

I use the method to show how spacing between births has changed with
the introduction prenatal sex determination and how birth spacing plays 
an important role in our understanding of sex selection decisions and 
son preferences.

% [Why India?]
I apply the method to birth histories from India's National Family and Health 
Surveys for Hindu women covering the period 1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period \citep{Yi1993,park95}.
}
On the other hand, recent research suggests, however, that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}

% [EXPLAIN RESULTS HERE]
% Main findings!
There are three main findings.

@@
